---
title: "regfusionr: Registration Fusion in R"
author: "Tim Schäfer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{regfusionr: Registration Fusion in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE    # These examples require FreeSurfer, which is not available on CRAN check machines.
)
```

## Overview

The `regfusionr` package implements [Wu et al. (2018)'s registration fusion methods](https://onlinelibrary.wiley.com/doi/full/10.1002/hbm.24213) in R. It provides accurate nonlinear mapping between three standard coordinate systems commonly used in neuroimaging:

- **FreeSurfer fsaverage** surface space (MNI305)
- **FSL MNI152** volume space
- **SPM Colin27** volume space

If you use this package, please cite the original paper:

> Wu J, Ngo GH, Greve DN, Li J, He T, Fischl B, Eickhoff SB, Yeo BTT. **Accurate nonlinear mapping between MNI volumetric and FreeSurfer surface coordinate systems**, *Human Brain Mapping* 39:3793–3808, 2018.

## Prerequisites

All examples in this vignette require a working [FreeSurfer](https://surfer.nmr.mgh.harvard.edu/) installation. The `FREESURFER_HOME` environment variable must be set to the FreeSurfer installation directory (e.g., `/Applications/freesurfer/7.4.0`), or you can pass the path explicitly via the `fs_home` parameter.

If you have properly installed and configured FreeSurfer, this should already be the case. You can check that the `FREESURFER_HOME` environment variable is set correctly:

```{r}
Sys.getenv("FREESURFER_HOME")
```

The `vol_to_fsaverage()` function additionally requires the `oce` package for trilinear interpolation. The `fsaverage_to_vol()` function requires `fsbrain` and `haze`. These are optional dependencies and can be installed as needed.

```{r}
library("regfusionr")
```

## Mapping Volume Coordinates to fsaverage Surface

### MNI152 coordinates to fsaverage

The most common use case: you have a peak coordinate from an fMRI paper in MNI152 space, and you want to know where it is on the fsaverage surface.

```{r}
# Map a single MNI152 coordinate to fsaverage.
mni_ras = c(60.0, 0.0, 10.0)
res = mni152_coords_to_fsaverage(mni_ras, surface = "white")
res$fsaverage_vertices   # The fsaverage vertex index
res$hemi                 # Which hemisphere
res$fsaverage_coords     # The MNI305 (fsaverage) RAS coordinates
```

```{r}
# Map multiple coordinates at once.
mni_ras_matrix = matrix(c(60, 0, 10, 0, 0, 0), ncol = 3, byrow = TRUE)
res = mni152_coords_to_fsaverage(mni_ras_matrix, surface = "white")
# Coordinates outside cortex return NaN.
res$fsaverage_vertices   # e.g., c(9092, NaN)
```

### Colin27 coordinates to fsaverage

```{r}
c27_ras = c(60.0, 0.0, 10.0)
res = colin27_coords_to_fsaverage(c27_ras, surface = "white")
res$fsaverage_vertices
```

## Mapping fsaverage Vertices to Volume Coordinates

### fsaverage vertices to MNI152

```{r}
# Map fsaverage vertex 9092 (right hemisphere) to MNI152 RAS coordinates.
vertices = c(9092L)
hemis = c("rh")
mni_coords = fsaverage_vertices_to_mni152_coords(vertices, hemis)
```

### fsaverage vertices to Colin27

```{r}
vertices = c(9092L)
hemis = c("rh")
c27_coords = fsaverage_vertices_to_colin27_coords(vertices, hemis)
```

### Combining both hemispheres

```{r}
vertices = c(9092L, 100L)      # vertex 9092 (rh) and vertex 100 (lh)
hemis = c("rh", "lh")
coords = fsaverage_vertices_to_vol_coords(vertices, hemis,
  rf_type = "RF_ANTs", template_type = "MNI152_orig")
# coords is a 2x3 matrix of MNI152 RAS coordinates.
```

## Cross-Template Volume-to-Volume Mapping

Convenience functions let you map directly between MNI152 and Colin27 space via fsaverage.

### MNI152 to Colin27

```{r}
mni_ras = c(60.0, 0.0, 10.0)
c27_coords = mni152_coords_to_colin27_coords(mni_ras, surface = "white")
```

### Colin27 to MNI152

```{r}
c27_ras = c(60.0, 0.0, 10.0)
mni_coords = colin27_coords_to_mni152_coords(c27_ras, surface = "white")
```

## Projecting Volume Data to the fsaverage Surface

Instead of mapping individual coordinates, you can project an entire 3D or 4D volume to per-vertex surface data.

```{r}
# Project an MNI152 volume file to fsaverage surface space.
# The result is stored in R memory (out_dir = NULL).
input_file = system.file("extdata", "testdata",
  "MNI_probMap_ants.central_sulc.nii.gz", package = "regfusionr")
res = vol_to_fsaverage(input_file, template_type = 'MNI152_orig',
  rf_type = 'RF_ANTs', out_dir = NULL)
# res$lh: per-vertex data for the left hemisphere (numeric vector, length 163842)
# res$rh: per-vertex data for the right hemisphere

# To write output files instead, specify an out_dir.
out_dir = tempfile("regfusionr_output_")
res = vol_to_fsaverage(input_file, template_type = 'MNI152_orig',
  rf_type = 'RF_ANTs', out_type = 'curv', out_dir = out_dir)
# res$lh and res$rh now contain file paths to the written curv files.
```

## Projecting Surface Data to a Volume

Convert per-vertex fsaverage data (e.g., cortical thickness maps) to an MNI152 or Colin27 volume.

```{r}
# Create synthetic per-vertex data (fsaverage: 163842 vertices per hemisphere).
lh_data = rnorm(163842L, 3.0, 0.2)
rh_data = rnorm(163842L, 3.0, 0.2)

# Project to MNI152 volume space using linear interpolation.
# The result is kept in memory only (out_dir = NULL).
res = fsaverage_to_vol(lh_data, rh_data, target_space = "FSL_MNI152",
  interp = "linear", out_dir = NULL)
# res$projected:   fs.volume instance with a 256x256x256 data array.
# res$projected_seg: same, but right-hemisphere values incremented by 1000
#                    (useful for distinguishing hemispheres in labels).
```

For categorical label data, use `interp = "nearest"`:

```{r}
# Create synthetic label data per vertex.
lh_labels = sample.int(20, size = 163842L, replace = TRUE)
rh_labels = sample.int(20, size = 163842L, replace = TRUE)
res = fsaverage_to_vol(lh_labels, rh_labels, target_space = "FSL_MNI152",
  interp = "nearest", out_dir = NULL)
```

## MNI305 to MNI152/Colin27 via Closest Vertex

If you have MNI305 (fsaverage) coordinates and want MNI152 or Colin27 coordinates, you can find the closest fsaverage vertex and map it.

```{r}
# Get MNI305 coords of fsaverage vertex 9092 from the orig surface.
fsaverage_home = file.path(Sys.getenv("FREESURFER_HOME"), "subjects")
rh_orig = freesurferformats::read.fs.surface(
  file.path(fsaverage_home, "fsaverage", "surf", "rh.orig"))
mni305_pt = rh_orig$vertices[9092L, ]

# Map to MNI152.
mni152_coords = mni305_coords_to_mni152_coords(mni305_pt)

# Map to Colin27.
c27_coords = mni305_coords_to_colin27_coords(mni305_pt)
```

## Linear Method (for Comparison)

The package also includes the linear 4×4 matrix method from the FreeSurfer documentation, which you can use to compare accuracy with the nonlinear regfusion approach.

```{r}
# This function does NOT require FreeSurfer and CAN be run anywhere.
fsaverage_verts = matrix(c(10.0, 20.0, 30.0, 0.0, 0.0, 0.0), ncol = 3, byrow = TRUE)
mni152_linear = linear_fsaverage_coords_to_MNI152_coords(fsaverage_verts)
```

## Choosing RF-ANTs vs RF-M3Z

The package supports two registration fusion approaches:

- **RF-ANTs** (default): Recommended when ANTs or similar tools were used for registration. Use with `MNI152_orig` or `Colin27_orig` templates.
- **RF-M3Z**: Recommended when FreeSurfer was used for registration. Use with `MNI152_norm` or `Colin27_norm` templates.

For the coordinate-mapping functions (`mni152_coords_to_fsaverage`, `colin27_coords_to_fsaverage`), RF-M3Z uses a fallback `.txt`-based nearest-vertex method. The RF-ANTs MNI152 combination uses a faster pre-computed vertex index volume.

## Session Info

```{r}
sessionInfo()
```
